\texttt{maoud}: a Python package for Simulating Generalized Fading Channels

### 2018-03-23 paper/manuscript.tex
updates

\documentclass{sbrt2013port}
%\usepackage[brazil]{babel}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color}
%\usepackage{placeins}
\usepackage{bm}
%\usepackage{cite}
%\usepackage{stfloats}
%\usepackage{float}
\usepackage{hyperref}
%\usepackage{cite}
\usepackage{tabularx,colortbl}

\begin{document}

\title{Simulating Generalized Fading Channels in Python}

\author{José Vinícius de M. Cardoso$^1$,   Paulo Ribeiro Lins Júnior$^2$, Wamberto Queiroz$^1$ e Marcelo Sampaio de Alencar$^3$
\thanks{\noindent Instituto de Estudos Avançados em Comunicações -- Iecom; $^1$Universidade Federal de Campina Grande -- UFCG; $^2$Grupo de Pesquisa em Comunicações e Processamento de Informação -- GComPI/Instituto Federal de Educação, Ciência e Tecnologia da Paraíba -- IFPB, Campus Campina Grande -- PB, $^3$Universidade Federal da Bahia -- UFBA, emails: $^1$\{josevinicius, wamberto\}@iecom.org.br, $^2$paulo.lins@ifpb.edu.br e $^3$malencar@iecom.org.br.}}

@@

\markboth{XXXVI SIMPÓSIO BRASILEIRO DE TELECOMUNICAÇÕES E PROCESSAMENTO DE SINAIS - SBrT'2018, 16-19 DE SETEMBRO DE 2018, CAMPINA GRANDE, PB} {XXXVI SIMPÓSIO BRASILEIRO DE TELECOMUNICAÇÕES E PROCESSAMENTO DE SINAIS - SBrT'2018, 16-19 DE SETEMBRO DE 2018, CAMPINA GRANDE, PB}

\begin{abstract}
    We present a novel and well tested Python-based library for simulating and computing
    generalized fading channels, considering $\alpha\text{-}\mu$, $\kappa\text{-}\mu$, and
$\eta\text{-}\mu$ distributions. We describe the
    applicability of this library using a few communications channels scenarios
    impaired by generalized fading, namely, spectrum sensing and bit error
    rate computation. The development of the library is open source and its code, along
    with examples, are available at
    \texttt{http://github.com/mirca/maoud}.
\end{abstract}

\begin{keywords}
Generalized Fading, Simulation, Python
\end{keywords}

\section{Introduction}

The study of modern wireless communications systems heavily relies on fading channel simulation.
By fading channel simulation, we refer to the generation of samples from a probability distribution
that resembles the effects and impairments caused by real communications fading channels on the
transmitted signal.

Although accurate and precise distributions for generalized fading have been
estabilished in the literature, such as $\alpha\text{-}\mu$~\cite{yacoub.2007.1},
$\kappa\text{-}\mu$~\cite{yacoub.2007.2} and $\eta\text{-}\mu$~\cite{yacoub.2007.2},
the generation of samples following these distributions is usually
a time-consuming task.

In~\cite{cogliatti2012}, the authors built an efficient
algorithm, based on the rejection method, for generation of samples from those
distributions. However, there are neither open nor
closed source implementations available to the scientific community.

In this paper, we present an open source Python package
for generation of samples following the $\alpha\text{-}\mu$, $\kappa\text{-}\mu$, and
$\eta\text{-}\mu$ distributions. The usefullness of ths package is illustrated
through well known examples involving spectrum sensing and bit error rate (BER) computation.

\section{Rejection Sampling}\label{rejec}

The algorithm implemented in this work, originally proposed in~\cite{cogliatti2012},
is based on the rejection method, which is one of the most used for the generation
of samples of a probability distribution.

We briefly describe the algorithm as follows. Consider that we would like to simulate a
random variable $x$ with a probability density $f$. The basic idea is to find
an alternative probability density $g$, from which there exists an efficient procedure
to generate samples, e.g., inverse methods. Additionally, it is required that $g$
majorizes $f$, i.e., we assume that there exists a real number $c$ such that $f(x)/g(x) < c$.

Basically, the steps of the rejection algorithm are~\cite{ross}:
\begin{enumerate}
	\item generate a sample $y$ from the probability density $g$;
    \item generate a sample $u$ from a uniform density with support $(0, 1)$;
	\item check $u \leq \frac{f(y)}{cg(y)}$; if true then accept $y$; otherwise, return to step 1.
\end{enumerate}

The implementation of this algorithm along with the code to efficiently generate
samples from the $\alpha$-$\mu$, $\eta$-$\mu$, and $\kappa$-$\mu$ distributions
is available at \url{http://github.com/mirca/maoud}.

In order to validate the implementation, we apply the routines to
two use cases in wireless communications: spectrum sensing and bit error rate computation.

\section{Spectrum Sensing in Complex Generalized Fading Channels}~\label{sensing}

@@

From a probabilistic point of view, the spectrum sensing problem may be framed as
a decision theory problem, as follows
\begin{align}
    H_0:  \bm{y} &= \bm{w},\\
    H_1:  \bm{y} &= h\bm{s} + \bm{w},
\end{align}
\noindent in which $\bm{y} \in \mathbb{C}^{n\times 1}$ is the received vector signal,
$\bm{w} \in \mathbb{C}^{n\times 1}$ is complex Gaussian noise process with zero mean
vector and covariance matrix given as $\sigma^2\bm{I}_n$, and $h$ is the channel gain.

In~\cite{cardoso2017}, the authors have shown that the cumulative distribution function (cdf) of the
energy statistic $\tilde{y} \triangleq \bm{y}^{\dagger}\bm{y}$ conditioned on the knowledge of $h$,
in case that $\bm{s}$ is an $M$-PSK signal such that every symbol has the same probability of occurrence,
$\mathbb{P}(s_n = s) = \frac{1}{M}$, is given as
\begin{align}
    P(\tilde{y} | h, H_1) = 1 - \mathrm{Q}_{n}\left(\sqrt{\dfrac{2n|h|^2E_s}{\sigma^2}}, \sqrt{\dfrac{2\tilde{y}}{\sigma^2}}\right),
\end{align}
in which $\mathrm{Q}_{n}$ is the Marcum-$\mathrm{Q}$ function and $E_s$ is the energy per symbol.

Recall that the energy detection rule can be expressed as
\begin{equation}
    d_\delta (\tilde{y}) = \mathbb{I}(\tilde{y} > \delta)
\end{equation}
in which $\mathbb{I}$ is the indicator function, $\delta$ is a strictly positive real
number known as energy threshold, and $d_\delta (\tilde{y}) = j,~j \in \{0,1\}$,
means that the detector has decided in favor of the hypothesis $H_j$.

As a result, the probabilities of false alarm and miss detection can
be written as
\begin{align}
    \mathbb{P}\left(d_\delta(\tilde{y}) = 1 | H_0\right) &=
        1 -  P(\delta | H_0) = 1 - \gamma\left(n, \frac{\delta}{\sigma^2}\right),\label{eq:pf} \\
    \mathbb{P}\left(d_\delta(\tilde{y}) = 0 | H_1\right) & = \int_{-\infty}^{+\infty} P(\delta | h, H_1)p(h)\;\mathrm{d}h,\label{eq:pd}
\end{align}
in which $\gamma$ is the regularized lower incomplete Gamma function, $p(h)$ is the pdf of the fading,
and $\delta$ is the energy detection threshold.

The performance of detection schemes can be measured by computing the Receiver Operating
Characteristic (ROC), which consists in varying $\delta$ and computing the pairs of
probability of false alarm and miss detection, as illustrated in Fig.~\ref{fig:spectrum-sensing}, in which the solid curve represents the theorectical probabilities as stated on
(\ref{eq:pf}) and (\ref{eq:pd}) for different values of
$\delta$, whereas bullets represent Monte Carlo simulations acquired with $10^6$ realizations.
The input signal $\bm{s}$ consists in a vector of length $n=25$ in which each entry represents
a symbol of a 64-PSK constellation. Symbols are assumed to be equiprobable and
the signal-to-noise ratio is set to $5$ dB.

\begin{figure}[!htb]
    \centering
    \includegraphics{figures/spectrum_sensing}
    \caption{Receiver Operating Characteristic for the energy detector in $\alpha$-$\mu$
             channel with $\alpha=2$ and $\mu=1$.
             }
    \label{fig:spectrum-sensing}
\end{figure}

\section{Bit Error Rate in $\alpha$-$\mu$ Fading}~\label{ber}

Consider the system
\begin{align}
    y = hs + w
\end{align}
in which $s \in \{0, a\}$, $a \in \mathbb{R}_{+}$, is a transmitted signal,
$h$ is an $\alpha$-$\mu$ random variable and $\bm{w}$ is a Gaussian random variable
with zero mean vector and variance $\sigma^2$, and $y$ is the received signal.

Assume that the binary symbols are equiprobable, then the probability of one bit error
is given as
\begin{align}
    p_{e} = \dfrac{1}{2}\left(\mathbb{P}\left(\hat{y} = 0 | s = a\right)
                            + \mathbb{P}\left(\hat{y} = 1 | s = 0\right)\right).
\end{align}

Further, assume that the decoded bit $\hat{y}$ is estimated using the minimum distance decoding
rule, i.e.,
\begin{align}
    \hat{y} = \mathbb{I}(|y - a|^2 < |y|^2)
\end{align}
therefore
\begin{align}
    \mathbb{P}\left(\hat{y} = 1 | s = 0\right) & = \mathbb{P}\left(|w - a|^2 - |w|^2 < 0\right) = 1 - \Phi\left(\frac{a}{2\sigma}\right)
\end{align}
and likewise
\begin{align}
    \mathbb{P}\left(\hat{y} = 0 | s = a\right) & = \mathbb{P}\left(ha + w < \frac{a}{2}\right) = \mathbb{E}_h\left(\Phi\left(\frac{a(1 - 2h)}{2\sigma}\right)\right).
\end{align}

In Fig.~\ref{fig:ber}, we show the expression of the BER as a function of the signal-to-noise ratio,
for the system under consideration with $\alpha=2$ and $\mu = 5$.
As we can see, the simulation presents an excellent agreement against theorectical results.
\begin{figure}[!htb]
    \centering
    \includegraphics{figures/ber}
    \caption{BER as a function of the SNR for $\alpha=2$ and $\mu = 5$}
    \label{fig:ber}
\end{figure}

\section{Conclusions}\label{conclusions}

This paper presents a novel Python-based library for generalized fading simulation.
The implementation is tested and validated in two use cases, namely,
spectrum sensing and BER computation.

The computational results show that the implementation presents a
satisfactory agreement when compared with the theorectical formulation.

We expect that the proposed library, which is open source,
will be usefull to researchers, professors, and students of
the field of wireless communications.

Finally, all the simulations were done in a personal laptop,
and the generation of $10^6$ samples from any of the generalized fading
distributions took on average 10 seconds.

\bibliographystyle{IEEEtran}
\bibliography{manuscript}

>>>>

\documentclass[conference, 10pt]{IEEEtran}
\usepackage[cmex10]{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color}
\usepackage{placeins}
\usepackage{bm}
\usepackage{cite}
%\usepackage{stfloats}
\usepackage{float}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{tabularx,colortbl}

\begin{document}
\title{\texttt{maoud}: a Python package for Simulating Generalized Fading Channels}

\author{\IEEEauthorblockN{Jos\'e~V.~de~M.~Cardoso,~Wamberto~J.~L.~Queiroz,}
        \IEEEauthorblockN{~Paulo~R.~Lins~J\'unior,~and~Marcelo~S.~Alencar,~\textit{IEEE Senior Member}}
\IEEEauthorblockA{Universidade Federal de Campina Grande\\
Instituto Federal de Educa\c c\~ao, Ciencia, e Tecnologia da Para\'iba\\
Campina Grande, Para\'iba, Brasil\\
\{josevinicius,paulo,wamberto,malencar\}@iecom.org.br\\}
}

@@

\begin{abstract}
    We present a well tested Python-based library for simulating and computing
    generalized fading channels, named \texttt{maoud}. We describe the
    applicability of \texttt{maoud} using examples in scenarios of communications
    channels impaired by generalized fading, namely: spectrum sensing, bit error
    rate computation, and fading estimation. For the latter, we develop an iterative
    algorithm using the Majorization-Minimization framework, which allows reliable
    estimation of the fading parameter. The development of \texttt{maoud} is open
    source and its code along with examples are avaliable at
    \texttt{http://github.com/mirca/maoud}.
\end{abstract}

\IEEEpeerreviewmaketitle
\section{Introduction}

\subsection*{Notation}
Scalars and random variables are denoted as \textit{italic}, small-case letters
\textit{e.g.} $x$; sets and events are denoted as \textit{italic}, capital
letters \textit{e.g.}, $A$; vectors and random vectors are denoted as
\textit{italic}, boldface, small-case letters \textit{e.g.} $\bm{x}$. The $n-$th
component of a vector $\bm{x}$ is denoted as $x_n$. A complex vector of length
$n$ is defined as $\bm{x} \in \mathbb{C}^{n\times 1}$. All vectors are column
vectors. Matrices are denoted as \textit{italic}, boldface, capital letters as
in $\bm{X}$; the identity matrix of order $n$ is denoted as $\bm{I}_n$. We
define a discrete-time circularly symmetric Gaussian process $\bm{z}$ as any
collection of random variables $\bm{z}~=~\bm{x}~+~j\bm{y}$,
$j \triangleq \sqrt{-1}$, such that $\bm{x}$ and $\bm{y}$ are i.i.d. jointly
Gaussian, with zero mean vector and covariance matrix given by
$\mathbb{E}\left[\bm{z}\bm{z}^{\dagger}\right]$, in which $\bm{z}^\dagger$ means
the conjugate transpose of $\bm{z}$. The expectation value w.r.t. to the
probability distribution of a random variable $x$ is denoted as $\mathbb{E}_x$.
The probability of an event $A$ is denoted as $\mathbb{P}(A)$. The indicator
function is denoted as $\mathbb{I}(\cdot)$, it evaluates to one if its argument
is true and zero otherwise. For any given two real functions $f$ and $g$ defined
on the same domain $D$, $f \cong g$ means that there exist a constant $c$ such that
$f(\bm{x}) = g(\bm{x}) + c$, $\forall~ \bm{x} \in D$.

\section{The Acceptance-Rejection sampler in log-space}

\section{Majorization-Minimization Algorithms}

\section{Examples}
\subsection{Spectrum Sensing in Complex Generalized Fading Channels}

@@

From a probabilistic point of view, the spectrum sensing problem may be framed as
a decision theory problem, as follows
\begin{align}
    H_0:~& \bm{y} = \bm{w},\\
    H_1:~& \bm{y} = h\bm{s} + \bm{w},
\end{align}
in which $\bm{y} \in \mathbb{C}^{n\times 1}$ is the decoded received vector signal,
$\bm{w} \in \mathbb{C}^{n\times 1}$ is complex Gaussian noise process with zero mean
vector and covariance matrix given as $\sigma^2\bm{I}_n$, and $h$ is the channel gain.

In~\cite{cardoso2017}, the authors have shown that the probability distribution of the
energy statistic $\tilde{y} \triangleq \bm{y}^{\dagger}\bm{y}$ conditioned on the knowledge of $h$,
in case that $\bm{s}$ is an $M$-PSK signal such that every symbol has the same probability of occurrence,
$\mathbb{P}(s_n = s) = \frac{1}{M}$, is given as
\begin{align}
    p(\tilde{y} | h, H_1) = 1 - Q_{n}\left(\sqrt{\dfrac{2n|h|^2E_s}{\sigma^2}}, \sqrt{\dfrac{2\tilde{y}}{\sigma^2}}\right),
\end{align}
in which $Q_{n}$ is the Marcum-$Q$ function and $E_s$ is the energy per symbol.

The pdf of $\tilde{y}$ can be written using the Law of Total Expectation
\begin{align}
    p(\tilde{y} | H_1) = \mathbb{E}_{h}\left[p(\tilde{y} | h, H_{1})\right]
                 = \int_{-\infty}^{+\infty} p(\tilde{y} | h, H_1)p(h)\;\mathrm{d}h.
\end{align}

Recall that the energy detection rule can be expressed as
\begin{equation}
    d_\delta (\tilde{y}) = \mathbb{I}(\tilde{y} > \delta)
\right.
\end{equation}
in which $\delta$ is a strictly positive real number known as energy threshold,
and $d_\delta (\tilde{y}) = j,~j \in \{0,1\}$, means that the detector has decided
in favor of the hypothesis $H_j$.

As a result, the probabilities of false alarm and miss detection can
be written as
\begin{align}
    p_f &\triangleq \mathbb{P}\left(d_\delta(\tilde{y}) = 1 | H_0\right) = 1 -  p(\delta | H_0),\label{eq:pf} \\
    p_d &\triangleq \mathbb{P}\left(d_\delta(\tilde{y}) = 0 | H_1\right) = \mathbb{E}_{h}\left[p(\delta, h | H_1)\right],
\label{eq:pd}
\end{align}

\subsection{Parameter Estimation in Nakagami-$m$ fading}
The Nakagami-$m$ density is given as
\begin{align}
    p(\bm{h})& = \prod_{i=1}^{n}\dfrac{2m^m}{\Gamma(m)\Omega^{m}}h_i^{2m - 1}
              \exp\left(-\dfrac{mh_i^2}{\Omega}\right) \nonumber \\
          = & \left(\dfrac{2m^m}{\Gamma(m)\Omega^{m}}\right)^{n}
          \exp\left(-\dfrac{m\sum_{i=1}^{n}h_i^2}{\Omega}\right) \prod_{i=1}^{n}h_i^{2m - 1}
\end{align}
And the log likelihood function (up to an additive constant) is given as
\begin{align}
\log p(\bm{h}) \cong &~n\left(m\left(\log m - \log\Omega\right) - \log\Gamma(m)\right)\nonumber
    \\ & -m\left(\dfrac{\sum_{i=1}^{n}h_i^2}{\Omega} - 2\sum_{i=1}^{n}\log h_i\right)
    \label{eq:loglike}
\end{align}

A direct maximum likelihood estimator (MLE) for (\ref{eq:loglike}) has been
investigated to be infeasible~\cite{cheng2001}.

Therefore, we use a Majorization-Minimization algorithm to find smooth
and easy to optimize lower bounds for $\log p(\bm{h})$. More precisely,
we need to find lower bounds for $m\log m$ and $-\log \Gamma(m)$ for $m \geq \frac{1}{2}$.
The former function is convex for $m \geq 0$, hence it can be lower bounded
by its first order Taylor series as
\begin{align}
    m \log m \geq m(1 + \log m_t) - m_t,
    \label{eq:lower-bound-mlogm}
\end{align}
The function $-\log \Gamma(m)$ is concave, therefore, it can be lower bounded
by its second order Taylor series expansion as
\begin{align}
    -\log \Gamma(m) \geq& - \log \Gamma(m_t) - \psi(m_t) (m - m_t)\nonumber\\
                        & - \frac{\psi'\left(\frac{1}{2}\right)}{2}(m - m_t) ^ 2,
                        ~m \geq \frac{1}{2},~m_t \geq \frac{1}{2},
    \label{eq:lower-bound-negloggamma}
\end{align}
in which $\psi(x) = \dfrac{\Gamma'(x)}{\Gamma(x)}$ is known as the digamma function.
For both inequalities presented above, equality is achieved at $m = m_t$.

Substituting (\ref{eq:lower-bound-mlogm}) and (\ref{eq:lower-bound-negloggamma}) into
(\ref{eq:loglike}), a lower bound for $\log p(\bm{h})$ is obtained in~(\ref{eq:surrogate}).
Due to the simple form of $g(m | m_t)$, its maximizer can be found in closed form, and
an updating rule for the MLE can be written as
\begin{align}
    m_{t+1} = 1 + \dfrac{1}{\psi'(\frac{1}{2})}&\left(1 + \log \frac{m_t}{\Omega} - \psi(m_t)
    + \right.\nonumber\\
    &+ \left.\dfrac{1}{n} \left(2\sum_{i=1}^{n}\log h_i - \frac{\sum_{i=1}^{n}h_i^2}{\Omega}\right)\right)
\end{align}

Additionally,
note that $m_{t+1}$ is an estimator of the true parameter value $m$. Its expected value,
conditioned on the knowledge of $m_t$, is given as
\begin{align}
 \mathbb{E}(m_{t+1} | m_t) = 1 + \dfrac{1}{\psi'(\frac{1}{2})}\left(\log \frac{m_t}{\Omega} - \psi(m_t)
    + \dfrac{2}{n}\sum_{i=1}^{n}\mathbb{E}(\log h_i)\right)
\end{align}

Its variance, also conditioned on the knowledge of $m_t$, is given as
\begin{align}
    \mathrm{var}(m_{t+1} | m_t) = \left(\dfrac{1}{n\psi'(\frac{1}{2})}\right)^2
    \mathrm{var}\left(2\sum_{i=1}^{n}\log h_i
    - \frac{\sum_{i=1}^{n}h_i^2}{\Omega}\right).
\end{align}

The Cram\'er-Rao Lower Bound for any unbiased estimator of $m$, say $\hat{m}$, is given as~\cite{cheng2001}
\begin{align}
    \mathrm{var}(\hat{m}) = \dfrac{1}{n\left(\psi'(m) - \frac{1}{m}\right)}
\end{align}

\begin{figure*}[!htb]
\begin{align}
    g(m | m_t) = n\left(-m\log\Omega  + m(1 + \log m_t) - m_t
    -\log \Gamma(m_t) - \psi(m_t) (m - m_t) - \frac{\psi'\left(\frac{1}{2}\right)}{2}(m - m_t)^2\right)
    -m\left(\dfrac{\sum_{i=1}^{n}h_i^2}{\Omega} - 2\sum_{i=1}^{n}\log h_i\right)
    \label{eq:surrogate}
\end{align}
\end{figure*}

\subsection{BER in Complex $\alpha-\mu$ Fading}

Consider the system
\begin{align}
    \bm{y} = h\bm{s} + \bm{w}
\end{align}
in which $\bm{s} \in \mathbb{C}^{n\times 1}$ is a complex On-Off Keying (OOK) signal,
$h$ is a complex $\alpha-\mu$ random variable and $\bm{w}$ is a complex Gaussian process
with zero mean vector and covariance matrix equals $\sigma^2\bm{I}_n$, and $\bm{y}$ is
the received complex vector signal.

Assume that the OOK symbols are equiprobable and that there exist no interference
between the in-phase and quadrature components, then the probability of one bit error
is given as
\begin{align}
    p_{e} = \dfrac{1}{2}\left(\mathbb{P}\left(\hat{y}_i = 0 | s_{i} = 1\right)
                            + \mathbb{P}\left(\hat{y}_i = 1 | s_{i} = 0\right)\right).
\end{align}

Assume that the decoded vector $\bm{\hat{y}}$ is estimated using the minimum distance decoding
rule, i.e.,

\section{Conclusions}

\section*{Acknowledgement}
The authors would like to thank the Federal University of Campina Grande (UFCG)
and the Institute for Advanced Studies in Communications (Iecom) for supporting
this research.

\bibliographystyle{IEEEtran}
\bibliography{manuscript.bib}
